Abstraction Refinement for Emptiness Checking of Alternating Data Automata

### 2017-10-20 tacas/draft.tex
ok

\title{The Impact of Alternation}

@@

The language inclusion problem is recognized as being central to
verification of hardware, communication protocols and software
systems. A property is a specification of the correct executions of a
system, given as a set $\mathcal{P}$ of executions, and the
verification problem asks if the set $\mathcal{S}$ of executions of
the system under consideration is contained within $\mathcal{P}$.

This problem is at the core of widespread verification techniques,
such as automata-theoretic model checking \cite{VARDI94}, where
systems are specified as finite-state automata and properties defined
using Linear Temporal Logic \cite{Pnueli77}. However the bottleneck of
this and other related verification techniques is the intractability
of language inclusion (PSPACE-complete for finite-state automata on
finite words over finite alphabets).

@@

Although being convenient specification tools, automata over infinite
alphabets lack the decidability properties ensured by finite
alphabets. In general, when considering infinite data as part of the
input alphabet, language inclusion is undecidable and, even
complementation becomes impossible, for instance, for timed automata
\cite{AlurDill94} or finite-memory register automata
\cite{Kaminski94}. In some cases, one can recover theoretical
decidability, by restricting the number of variables (clocks) in timed
automata to one \cite{OuaknineWorrell04}, or forbidding relations
between current and past/future values, as with symbolic automata
\cite{symbTransd:POPL12}. In such cases, the emptiness problem for the
alternating versions becomes decidable \cite{Lasota05,DAntoniKW16}.

@@

The model presented in this paper stems from the language inclusion
problem considered in \cite{Tacas16}. There we provide a
semi-algorithm for inclusion of data languages, based on an
exponential determinization procedure and an abstraction refinement
loop using lazy predicate abstraction \cite{HJMS02}. In this work we
consider the full model of alternation and rely entirely on the
ability of SMT solvers to produce interpolants in the combined theory
of booleans and data. Since determinisation is not needed and
complementation is possible in linear time, the bulk of the work is
carried out by the decision procedure.

@@

A \emph{signature} $\sig = (\ssorts{\sig},\sfuns{\sig})$ consists of a
set $\ssorts{\sig}$ of \emph{sort symbols} and a set $\sfuns{\sig}$ of
sorted \emph{function symbols}. To simplify the presentation, we
assume w.l.o.g. that $\ssorts{\sig} = \set{\Data,\Bool}$\footnote{The
  generalization to more than two sorts is without difficulty, but
  would unnecessarily clutter the technical presentation.} and each
function symbol $f \in \sfuns{\sig}$ has $\#(f) \geq 0$ arguments of
sort $\Data$ and return value $\sigma(f) \in \ssorts{\sig}$. If
$\#(f)=0$ then $f$ is a \emph{constant}. We consider the constants
$\top$ and $\bot$ of sort $\Bool$.

We consider an infinite countable set of \emph{variables} $\vars$,
where each $x \in \vars$ has an associated sort $\sigma(x)$. A
\emph{term} $t$ of sort $\sigma(t)=S$ is a variable $x \in \vars$
where $\sigma(x)=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
$t_1,\ldots,t_{\#(f)}$ are terms of sort $\Data$ and $\sigma(f)=S$. An
\emph{atom} is a term of sort $\Bool$ or an equality $t \teq s$
between two terms of sort $\Data$. A \emph{formula} is an
existentially quantified combination of atoms using disjunction
$\vee$, conjunction $\wedge$ and negation $\neg$ and we write $\phi
\rightarrow \psi$ for $\neg\phi \vee \psi$.

@@

An \emph{alternating data automaton} (ADA or automaton in the sequel)
is a tuple $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$,
where: \begin{compactitem}
%
\item $\vec{x} \subset \vars$ is a finite set of variables of sort
  $\Data$,
%
\item $Q \subset \vars$ is a finite set of variables of sort $\Bool$
  (\emph{states}),
%
\item $\iota \in \Form^+(Q,\emptyset)$ is the \emph{initial
  configuration},
%
\item $F \subseteq Q$ is a set of \emph{final states}, and
%
\item $\Delta : Q \times \Sigma \rightarrow
  \Form^+(Q,\overline{\vec{x}}\cup\vec{x})$ is a \emph{transition
    function},
\end{compactitem}
where $\overline{\vec{x}}=\{\overline{x} \mid x \in \vec{x}\}$. In
each formula $\Delta(q,a)$ describing a transition rule, the variables
$\overline{\vec{x}}$ track the previous and $\vec{x}$ the current
values of the variables of $\A$. Observe that the initial values of
the variables are left unconstrained, as the initial configuration
does not contain free data variables. The size of $\A$ is defined as
$\len{\A} = \len{\iota} + \sum_{(q,a) \in
  Q\times\Sigma}\len{\Delta(q,a)}$.

>>>>

\title{Abstraction Refinement for Emptiness Checking of Alternating
  Data Automata}

@@

The language inclusion problem is recognized as being central to
verification of hardware, communication protocols and software
systems. A property is a specification of the correct executions of a
system, given as a set $\mathcal{P}$ of executions, and the
verification problem asks if the set $\mathcal{S}$ of executions of
the system under consideration is contained within $\mathcal{P}$.
This problem is at the core of widespread verification techniques,
such as automata-theoretic model checking \cite{VARDI94}, where
systems are specified as finite-state automata and properties defined
using Linear Temporal Logic \cite{Pnueli77}. However the bottleneck of
this and other related verification techniques is the intractability
of language inclusion (PSPACE-complete for finite-state automata over
finite alphabets).

@@

Although being convenient specification tools, automata over infinite
alphabets lack the decidability properties ensured by finite
alphabets. In general, when considering infinite data as part of the
input alphabet, language inclusion is undecidable and, even
complementation becomes impossible, for instance, for timed automata
\cite{AlurDill94} or finite-memory register automata
\cite{Kaminski94}. One can recover theoretical decidability, by
restricting the number of variables (clocks) in timed automata to one
\cite{OuaknineWorrell04}, or forbidding relations between current and
past/future values, as with symbolic automata
\cite{symbTransd:POPL12}. In such cases, also the emptiness problem
for the alternating versions becomes decidable
\cite{Lasota05,DAntoniKW16}.

@@

The model presented in this paper stems from the language inclusion
problem considered in \cite{Tacas16}. There we provide a
semi-algorithm for inclusion of data languages, based on an
exponential determinization procedure and an abstraction refinement
loop using lazy predicate abstraction \cite{HJMS02}. In this work we
consider the full model of alternation and rely entirely on the
ability of SMT solvers to produce interpolants in the combined theory
of booleans and data. Since determinisation is not needed and
complementation is possible in linear time, the bulk of the work is
carried out by the solver.

@@

A \emph{signature} $\sig = (\ssorts{\sig},\sfuns{\sig})$ consists of a
set $\ssorts{\sig}$ of \emph{sort symbols} and a set $\sfuns{\sig}$ of
sorted \emph{function symbols}. To simplify the presentation, we
assume w.l.o.g. that $\ssorts{\sig} = \set{\Data,\Bool}$\footnote{The
  generalization to more than two sorts is without difficulty, but
  would unnecessarily clutter the technical presentation.} and each
function symbol $f \in \sfuns{\sig}$ has $\#(f) \geq 0$ arguments of
sort $\Data$ and return value $\sigma(f) \in \ssorts{\sig}$. If
$\#(f)=0$ then $f$ is a \emph{constant}. We consider constants $\top$
and $\bot$ of sort $\Bool$.

Let $\vars$ be an infinite countable set of \emph{variables}, where
each $x \in \vars$ has an associated sort $\sigma(x)$. A \emph{term}
$t$ of sort $\sigma(t)=S$ is a variable $x \in \vars$ where
$\sigma(x)=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
$t_1,\ldots,t_{\#(f)}$ are terms of sort $\Data$ and $\sigma(f)=S$. An
\emph{atom} is a term of sort $\Bool$ or an equality $t \teq s$
between two terms of sort $\Data$. A \emph{formula} is an
existentially quantified combination of atoms using disjunction
$\vee$, conjunction $\wedge$ and negation $\neg$ and we write $\phi
\rightarrow \psi$ for $\neg\phi \vee \psi$.

@@

An \emph{alternating data automaton} (ADA or automaton in the sequel)
is a tuple $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$,
where: \begin{compactitem}
%
\item $\vec{x} \subset \vars$ is a finite set of variables of sort
  $\Data$,
%
\item $Q \subset \vars$ is a finite set of variables of sort $\Bool$
  (\emph{states}),
%
\item $\iota \in \Form^+(Q,\emptyset)$ is the \emph{initial
  configuration},
%
\item $F \subseteq Q$ is a set of \emph{final states}, and
%
\item $\Delta : Q \times \Sigma \rightarrow
  \Form^+(Q,\overline{\vec{x}}\cup\vec{x})$ is a \emph{transition
    function}, where $\overline{\vec{x}}$ denotes $\{\overline{x} \mid
  x \in \vec{x}\}$.
\end{compactitem}
In each formula $\Delta(q,a)$ describing a transition rule, the
variables $\overline{\vec{x}}$ track the previous and $\vec{x}$ the
current values of the variables of $\A$. Observe that the initial
values of the variables are left unconstrained, as the initial
configuration does not contain free data variables. The size of $\A$
is defined as $\len{\A} = \len{\iota} + \sum_{(q,a) \in
  Q\times\Sigma}\len{\Delta(q,a)}$.
